Towards Compiler-Agnostic Performance in Finite-Difference Codes

### 2015-09-15 docs/papers/shallow_performance/PSEA-2015_review_response.txt

> 6. para 1, page 4: is it one third per section?
Re-worded.

> 7. page 5, second transformation: remove second 'as'
Fixed.

> 8. para 3, page 6: 64, 128 ... should these be squared also on the figure too.
Fixed.

> 9. para 4, page 6: in the statement 'for the first two bars...' I
  would say that this is also true for the third bar also. This is gnu
  on the IvyBridge. Is that understood?
Text has been added to explain this.

> 11. para6, page 6: difference seems to be in italics.
Fixed.

> 12. para 2, page 8: could you reword this; "on Haswell loop fusion
 achieves more than 90% while on Ivy Bridge it is fully-fusing the
 first loop nest", I found it difficult to follow.
Re-worded.

> 13. Figure 4 and labeling the transformations: The fact that you
  have labeled the run instances 0-10 means its difficult to compare
  the labels with the bars (for me at least). Rather than reading the
  shading, I found my self counting them and I kept getting mixed up
  because the sixth bar corresponds to transformation 5. Further, in
  the text on page 9, there is reference to bar 6 which I assume is
  transformation 6 but bar 7... I understand why its been laid out
  this way but I think the transformations/vanilla could be labeled
  1-11?
We agree and have re-numbered the transformations from 1-11. We have
also added labels to the key optimisation steps in the Figure and
refer to the transformations by number in the text.

> 14. para 3, page 9: potentially change "operation of Figure 5 is
  fused" to "operation (lower loop in Figure 5) is fused."
Re-worded.

> 15. pare 1, page 10: in the sentence; "remain constant for the
> duration of each trip of the time-stepping loop" I've never heard
> that terminology before, is that standard? I would expect to hear
> something like "remain constant for the duration of each iteration
> of the time-stepping loop"
Re-worded.

### 2015-09-15 docs/papers/shallow_performance/shallow_perf.tex
parco paper: further tweaks for first reviewer

\begin{abstract}
In this paper we evaluate the performance implications of applying a
technique which we call \psykal\ to finite difference Ocean models. 
%
In \psykal, the code related to the underlying science is formally
separated from code related to parallelisation and single core
optimisations. This separation of concerns allows scientists to code
their science independently of the underlying hardware architecture
(thereby keeping a single code base) and for optimisation specialists
to be able to tailor the code for a particular machine independently
of the science code.
%
A finite difference shallow water benchmark optimised for cache based
architectures is taken as the starting point. A vanilla \psykal\ version
is written and the performance of the two compared. The optimisations
that were applied to the original benchmark (loop fusion {\it etc.}) are then
applied to the \psykal\ version as a set of code modifications to the
optimisation layer. Performance results are presented for the Cray,
Intel and GNU compilers on Intel Ivybridge and Haswell processors and
for the IBM compiler on Power8.
%
Results show that the combined set of code modifications obtain
performance that is within a few percent of the original code for all
compiler and architecture combinations on all tested problem
sizes. The only exception to this (other than where we see significant
speed-ups) is the Gnu compiler on Haswell for one problem size. Our
tests indicate that this may be due to immature support for that
architecture in the Gnu compiler -- no such problem is seen on the Ivy
Bridge system.
%%
Therefore, the \psykal\ approach can be used with negligable
performance loss and sometimes small performance gains compared to the
original optimised code, there is no single best hand optimised
implementation of the code for all the compilers tested, and the
flexibility in applying code modifications to the optimisation layer
in the PSyKAl version allows for a significant performance improvement
on the IBM Power8 compared with the original code.

@@

Before applying any code transformations, we benchmark the original
version of the code. We also benchmark the vanilla, unoptimised
version after it has been re-structured following the \psykal
approach. These two versions of the code effectively provide upper and
lower bounds, respectively, on the performance we expect to achieve.

Beginning with the vanilla \psykal\ version, we then manually apply a
series of (cumulative) code transformations while obeying the \psykal\ 
separation of concerns, {\it i.e.} optimisation is restricted to the
middle, {PS}y layer and leaves the kernel and algorithm layers
unchanged. The aim of these optimisations is to recover, as much as is
possible, the structure and thus performance of the original version
of the code. We began with two transformations that affect all of the
kernels. After this, we determined the transformations to perform by
comparing the timing profiles of the \psykal\ and original versions of
the code. We then tackled the sections resposible for the largest
slowdowns and iterated this process. While doing this we also made
heavy use of the optimisation reports produced by the compilers and
how these changed when moving from the original to the \psykal\ version
of the code.

The transformations we have performed are as follows:
\begin{enumerate}

@@

\end{enumerate}

As indicated above, transformations 10 and 11 were
suggested by the IBM compiler team. The second of these effectively
undoes transformation four by fissioning the first loop nest back
into three separate loop nests.

@@

We now examine the performance of the \psykal\ version of Shallow and
the code transformations that are required to reach the levels obtained
by the original.  We do this for the $256^{2}$ problem size since this
fits within last-level cache on all of the CPUs we are using here.
The performance of the \psykal\ version after each transformation has
been applied to the code is plotted in
Figure~\ref{FIG_opt_stages_256}. Results for a given compiler/CPU are
given relative to the performance of the original version of Shallow
with the same compiler/CPU.

@@

With the Gnu compiler (first and third clusters in
Figure~\ref{FIG_opt_stages_256}), the vanilla \psykal\ version of
Shallow only achieves ~5-6\% of the performance of the
original. Simply copying the code for each of the kernel subroutines
into the module from which they are called (module in-lining) has the
most dramatic effect on the performance of the compiled code; it now
achieves more than 80\% of the performance of the original. The
transformation(s) required to get more than 90\% then become
architecture dependent: on Haswell it is just loop fusion while on Ivy
Bridge we do not exceed 90\% until we have done all of the steps up to
and including fully-fusing (fusing the inner loops of) the first loop
nest. The dramatic performance improvement obtained from the
module-inlining transformation demonstrates that the Gnu compiler is
unable to optimise over separate files.

With the Intel compiler, the key transformations to recover
performance are different (second and fourth clusters in
Figure~\ref{FIG_opt_stages_256}).  Even the vanilla \psykal\ version of
Shallow achieves some 80\% of the performance of the original
version. Only two code transformations were required to increase this
to $\sim$100\%. The first of these was the fusion of the computational
loops. However, we found that performance suffered (bar 5) when this
was extended to include the loop required for field copies at the end
of a time-step (Figure~\ref{FIG_time_smooth_code}). Further
investigation of this slow-down revealed that the compiler was unable
to SIMD vectorise the newly-fused loop after it had in-lined the body
of the kernel called from within it. Once this in-lining was done by
hand (bar 6), the compiler was able to vectorise the loop and the
performance of the original version of Shallow was recovered on the
Haswell system. On Ivy Bridge the first loop nest also had to be fully
fused (bar 7) to recover $\sim100\%$ of the performance.

@@

The fifth cluster in Figure~\ref{FIG_opt_stages_256} shows the
evolution of the performance of the \psykal\ version of Shallow with
the Cray compiler. Again, this has significant differences from the
behaviour seen with the Gnu and Intel compilers. As with the Intel
compiler, the performance of the Vanilla \psykal\ version is fairly
good at 74\% of that of the original. In contrast to the other
compilers consisdered so far, the first significant transformation is
simply to specify the bounds of the arrays being used within the
middle layer using module variables (as opposed to specifying them as
assumed size). This tells the compiler that all of the arrays used in
the middle layer have the same extent and also that all of the
computational loops are over almost all elements of these arrays. This
simple change gives a \psykal\ version that achieves 88\% of the
performance of the original.

Subsequent transformations actually harm performance until the
field-copy operation of Figure~\ref{FIG_time_smooth_code} is fused
with the preceeding loop. The resulting \psykal\ version now achieves
94\% of the original performance. Two further steps are required to
match the latter. First, the first loop nest has to be fully fused
(both inner and outer loops fused) and second, the array/loop bounds
are passed as an argument to the middle layer. This is done in order
to indicate to the compiler that these bounds remain constant for the
duration of each trip of the time-stepping loop.

Moving to the IBM compiler on the Power 8 system (sixth cluster in
Figure~\ref{FIG_opt_stages_256}) we see that performance of the
vanilla \psykal\ version is only 5\% of the original and several
transformations are required to achieve comparable performance. As
with the Cray compiler, specifying array bounds explicitly improves
performance. Module-inlining also has a significant effect; doubling
the performance from 10 to 20\% of the original. However, fusing loop
nests actually harms performance, just as it did for the Cray
compiler. Kernel inlining takes performance to over 50\% and it is the
IBM-supplied transformations of loop unrolling and fissioning the
first loop ({\it i.e.} undoing some of transformations three and
seven) that finally recovers 96\% of the performance of the
original. We note that this final pair of transformations significantly
reduce the performance obtained with the Intel and Gnu compilers
although they do not much perturb that obtained by the Cray
compiler.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=120mm]{slowdown_summary}
\caption{Comparison of the performance of the best \psykal
version with that of the original version of the code. A negative value 
indicates that the \psykal\ version is faster than the original.}
\label{FIG_slowdown_summary}
\end{figure}

>>>>

\begin{abstract}
In this paper we evaluate the performance implications of applying a
technique which we call \psykal\ to finite difference Ocean models. 
%
In \psykal, the code related to the underlying science is formally
separated from code related to parallelisation and single core
optimisations. This separation of concerns allows scientists to code
their science independently of the underlying hardware architecture
(thereby keeping a single code base) and for optimisation specialists
to be able to tailor the code for a particular machine independently
of the science code.
%
A finite difference shallow water benchmark optimised for cache based
architectures is taken as the starting point. A vanilla \psykal\ version
is written and the performance of the two compared. The optimisations
that were applied to the original benchmark (loop fusion {\it etc.}) are then
applied to the \psykal\ version as a set of code modifications to the
optimisation layer. Performance results are presented for the Cray,
Intel and GNU compilers on Intel Ivybridge and Haswell processors and
for the IBM compiler on Power8.
%
Results show that the combined set of code modifications obtain
performance that is within a few percent of the original code for all
compiler and architecture combinations on all tested problem
sizes. The only exception to this (other than where we see significant
speed-ups) is the Gnu compiler on Haswell for one problem size. Our
tests indicate that this may be due to immature support for that
architecture in the Gnu compiler -- no such problem is seen on the Ivy
Bridge system.
%%
Therefore, the \psykal\ approach can be used with negligable
performance loss and sometimes small performance gains compared to the
original optimised code, there is no single best hand optimised
implementation of the code for all the compilers tested, and the
flexibility in applying code modifications to the optimisation layer
in the \psykal\ version allows for a significant performance improvement
on the IBM Power8 compared with the original code.

@@

Before applying any code transformations, we benchmark the original
version of the code. We also benchmark the vanilla, unoptimised
version after it has been re-structured following the \psykal\
approach. These two versions of the code effectively provide upper and
lower bounds, respectively, on the performance we expect to achieve.

Beginning with the vanilla \psykal\ version, we then manually apply a
series of (cumulative) code transformations while obeying the \psykal\ 
separation of concerns, {\it i.e.} optimisation is restricted to the
middle, {PS}y layer and leaves the kernel and algorithm layers
unchanged. The aim of these optimisations is to recover, as much as is
possible, the structure and thus performance of the original version
of the code. We began with two transformations that affect all of the
kernels. After this, we determined the transformations to perform by
comparing the timing profiles of the \psykal\ and original versions of
the code. We then tackled the sections resposible for the largest
slowdowns and iterated this process. While doing this we also made
heavy use of the optimisation reports produced by the compilers and
how these changed when moving from the original to the \psykal\ version
of the code. The transformations we have performed are as follows:

\newcounter{tnum}

\begin{list}{T\arabic{tnum}.}{\usecounter{tnum}\setlength{\rightmargin}{\leftmargin}}

@@

\end{list}

As indicated above, transformations T10 and T11 were
suggested by the IBM compiler team. The second of these effectively
undoes transformation four by fissioning the first loop nest back
into three separate loop nests.

@@

We now examine the performance of the \psykal\ version of Shallow and
the code transformations that are required to reach the levels
obtained by the original.  We do this for the $256^{2}$ problem size
since this fits within LLC on all of the CPUs we are using here.  The
performance of the \psykal\ version after each transformation has been
manually applied to the code is plotted in
Figure~\ref{FIG_opt_stages_256}. Results for a given compiler/CPU are
given relative to the performance of the original version of Shallow
with the same compiler/CPU.

@@

With the Gnu compiler (first and third clusters in
Figure~\ref{FIG_opt_stages_256}), the vanilla \psykal\ version of
Shallow only achieves ~5-6\% of the performance of the
original. Simply copying the code for each of the kernel subroutines
into the module from which they are called (module in-lining, T3) has the
most dramatic effect on the performance of the compiled code; it now
achieves more than 80\% of the performance of the original. The
transformation(s) required to get more than 90\% then become
architecture dependent: on Haswell it is just loop fusion (T4) while on Ivy
Bridge we do not exceed 90\% until we have done all of the steps up to
and including T8; fully fusing (fusing the inner loops of)
the first loop nest. The dramatic performance improvement obtained
from the module-inlining (T3) demonstrates that the Gnu
compiler is unable to optimise over separate source files.

With the Intel compiler, the key transformations to recover
performance are different (second and fourth clusters in
Figure~\ref{FIG_opt_stages_256}).  Even the vanilla \psykal\ version of
Shallow achieves some 80\% of the performance of the original
version. Only two code transformations were required to increase this
to $\sim$100\%. The first of these was T4, the fusion of the computational
loops. However, we found that performance suffered (T6) when this
was extended to include the loop required for field copies at the end
of a time-step (Figure~\ref{FIG_time_smooth_code}). Further
investigation of this slow-down revealed that the compiler was unable
to SIMD vectorise the newly-fused loop after it had in-lined the body
of the kernel called from within it. Once this in-lining was done by
hand (T7), the compiler was able to vectorise the loop and the
performance of the original version of Shallow was recovered on the
Haswell system. On Ivy Bridge the first loop nest also had to be fully
fused (T8) to recover $\sim$100\% of the performance.

@@

The fifth cluster in Figure~\ref{FIG_opt_stages_256} shows the
evolution of the performance of the \psykal\ version of Shallow with
the Cray compiler. Again, this has significant differences from the
behaviour seen with the Gnu and Intel compilers. As with the Intel
compiler, the performance of the Vanilla \psykal\ version, T1,  is fairly
good at 74\% of that of the original. In contrast to the other
compilers consisdered so far, the first significant transformation is
simply to specify the bounds of the arrays being used within the
middle layer using module variables (as opposed to specifying them as
assumed size). This tells the compiler that all of the arrays used in
the middle layer have the same extent and also that all of the
computational loops are over almost all elements of these arrays. This
simple change gives a \psykal\ version that achieves 88\% of the
performance of the original.

Subsequent transformations actually harm performance until T6 where
the field-copy operation (lower loop in
Figure~\ref{FIG_time_smooth_code}) is fused with the preceeding
loop. The resulting \psykal\ version now achieves 94\% of the original
performance. Two further steps are required to match the latter: the
first loop nest is fully fused (T8 - both inner and outer loops fused)
and second, the array/loop bounds are passed as an argument to the
middle layer (T9). This is done in order to indicate to the compiler
that these bounds remain constant for the duration of each iteration
of the time-stepping loop.

Moving to the IBM compiler on the Power 8 system (sixth cluster in
Figure~\ref{FIG_opt_stages_256}) we see that performance of the
vanilla \psykal\ version is only 5\% of the original and several
transformations are required to achieve comparable performance. As
with the Cray compiler, specifying array bounds explicitly, T2, improves
performance. Module-inlining, T3, also has a significant effect; doubling
the performance from 10 to 20\% of the original. However, fusing loop
nests actually harms performance, just as it did for the Cray
compiler. Kernel inlining (T7) takes performance to over 50\% and it is the
IBM-supplied transformations of T10 (loop unrolling) and T11 (fissioning of the
first loop) that finally recovers 96\% of the performance of the
original. (\textit{i.e.} we must undo some of transformations three and
seven). We note that this final pair of transformations significantly
reduce the performance obtained with the Intel and Gnu compilers
although they do not much perturb that obtained by the Cray
compiler.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=120mm]{slowdown_summary}
\caption{Comparison of the performance of the best \psykal\
version with that of the original version of the code. A negative value 
indicates that the \psykal\ version is faster than the original.}
\label{FIG_slowdown_summary}
\end{figure}
